BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-11-20 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
resubmitted version

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract: |
  \noindent
  **Premise:** Accessioning herbarium specimens is labor-intensive, yet remains vital for research in ecology, evolution, and conservation. As institutional support for herbaria declines, efficient tools are needed to streamline this process. BarnebyLives was developed to assist collectors by supplementing collection notes, verifying taxonomic data, conducting quality checks, generating labels, and submitting digital records.  
  **Methods and Results:** It integrates geospatial data from U.S. government sources to provide jurisdictional and site information, and checks taxonomic names using in-house spell checkers, IPNI author standards, and Kew’s Plants of the World Online. Optional features include generating Google Maps driving directions. The tool outputs data in tabular and spatial formats for review before producing LaTeX-based labels and shipping manifests.     
  **Conclusions:** BarnebyLives improves data accuracy, ensures up-to-date taxonomy, and significantly reduces the time and effort required to accession herbarium specimens.        
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output: 
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
  - \usepackage{endfloat}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{rotating}
---

@@

We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives. 
Given that the attributes which the package collects data on are tailored to the Western U.S. region, we do not expect local installs to exceed the size of ours. 
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
Unfortunately, we have not found Windows alternatives to pdfjam and pdftk, so we are unable to offer the final label-generating functionality on that operating system, but suspect Ubuntu subsystem for Windows may allow for integration of these tools.

## Functionality
BarnebyLives can be thought of as consisting of five main modules (Figure 1): spatial, taxonomic, formatting, manual review, and data exporting.

@@

While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate [@patten2024geographic].
We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to assume this responsibility in a relatively fast and streamlined format. 
While BarnebyLives may be adopted by herbarium curators, its original use was for natural resource management employees (for example, US Forest Service, heritage programs, natural history surveys) and graduate students in ecology and conservation (e.g., performing inventory and community ecology work), and we expect it to be most widely used in this realm. 
Beyond always having the data collected by BarnebyLives terminating in a label for a physical specimen, its utility may be extended to providing information for various biotic databases (e.g., heritage program databases) or modelling applications (e.g., modelling in community ecology) as a user sees fit.
By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches. 
LaTeX, a software system used for typesetting, was developed to allow users to focus on the content rather than the style of the rendered documents. 
Indeed, using the default LaTeX settings along with standard templates allows users to produce aesthetically pleasing results (Figure 4). 
However, LaTeX also offers a rich suite of customizable parameters that offer users a wide variety of ways to modify labels that are under-explored in the package.
Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn) project. 
While the templates in the package are quite simple, LaTeX also offers the ability to use custom fonts, to alter font weights and colors, alter line spacing, to include images (e.g. dot maps) and customize labels beyond what the default templates support.

@@

Accessioning often relies on the use of the Microsoft Office suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users. 
While other software packages often cost money, these are also free, and we recommend that users install an open-source PDF viewer such as Okular to review their rendered documents.

@@

# CONCLUSIONS
BarnebyLives is an R package that can be used to rapidly acquire relevant geographic and taxonomic data. 
It can also perform specialized spell checks and assorted curatorial tasks to produce both digital and analog data. 
The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Mac and Linux, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.

@@

# DATA AVAILABILITY STATEMENT
The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives). 
The package includes three real use-case vignettes (tutorials) available on a Github Pages site (https://sagesteppe.github.io/BarnebyLives/).
The first vignette *"Preparing to use BarnebyLives!"* shows how to set up an instance for a certain geographic area (domain). 
The next two vignettes *"BarnebyLives! Running pipeline"* showcases the usage of the package for processing data entered on a spreadsheet, and *"Printing herbarium labels and exporting a digital copy of data"* demonstrates how to export data in both digital and analog formats. 
*"Custom label templates"* shows how to customize labels in LaTeX, and *"Rendering a shipping manifest"* details how to produce a shipping manifest for gifting or transferring material to an herbarium. 
All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript.

>>>>

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
abstract: |
  \noindent
  **Premise:** Accessioning herbarium specimens is labor-intensive, yet remains vital for research in ecology, evolution, and conservation. As institutional support for herbaria declines, efficient tools are needed to streamline this process. BarnebyLives was developed to assist collectors by supplementing collection notes, verifying taxonomic data, conducting quality checks, generating labels, and submitting digital records.  
  **Methods and Results:** It integrates geospatial data from U.S. government sources to provide jurisdictional and site information, and checks taxonomic names using in-house spell checkers, IPNI author standards, and Kew’s Plants of the World Online. Optional features include generating Google Maps driving directions. The tool outputs data in tabular and spatial formats for review before producing LaTeX-based labels and shipping manifests.     
  **Conclusions:** BarnebyLives improves data accuracy, ensures up-to-date taxonomy, and significantly reduces the time and effort required to accession herbarium specimens.        
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output: 
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
  - \usepackage{endfloat}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{rotating}
---

@@

We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives. 
Given that the attributes which the package collects data on are tailored to the Western U.S. region, we do not expect local installs to exceed the size of ours. 
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
However, installation of the utilities on the three supported Operating systems (Debian Linux, MacOS, and Windows 10 and 11)  will require some minimal use of the command line to install software which are required for the label-generating functionality as detailed on the README page.

## Functionality
BarnebyLives can be thought of as consisting of five main modules (Figure 1): spatial, taxonomic, formatting, manual review, and data exporting. 
Core functions are detailed below, while all functions are documented on reference page https://sagesteppe.github.io/BarnebyLives/reference/index.html.

@@

While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate [@patten2024geographic].
We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to assume this responsibility in a relatively fast and streamlined format. 
While BarnebyLives may be adopted by herbarium curators, its original use was for natural resource management employees (for example, US Forest Service, heritage programs, natural history surveys) and graduate students in ecology and conservation (e.g., performing inventory and community ecology work), and we expect it to be most widely used in this realm. 
Beyond always having the data collected by BarnebyLives terminating in a label for a physical specimen, its utility may be extended to providing information for various biotic databases (e.g., heritage program databases) or modelling applications (e.g., modelling in community ecology) as a user sees fit.
By utilizing both R and LaTeX and having publicly available source code on GitHub, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches. 
LaTeX, a software system used for typesetting, was developed to allow users to focus on the content rather than the style of the rendered documents. 
Indeed, using the default LaTeX settings along with standard templates allows users to produce aesthetically pleasing results (Figure 4). 
However, LaTeX also offers a rich suite of customizable parameters that offer users a wide variety of ways to modify labels that are under-explored in the package.
Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn) project. 
While the templates in the package are quite simple, LaTeX also offers the ability to use custom fonts, to alter font weights and colors, alter line spacing, to include images (e.g. dot maps) and customize labels beyond what the default templates support.

@@

Accessioning often relies on the use of the Microsoft Office suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as RStudio, Positron, VisualStudio (VS) Code, now offer functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users. 
While other software packages often cost money, these are also free, and we recommend that users install an open-source PDF viewer such as Okular to review their rendered documents.

@@

# CONCLUSIONS
BarnebyLives is an R package that can be used to rapidly acquire relevant geographic and taxonomic data. 
It can also perform specialized spell checks and assorted curatorial tasks to produce both digital and analog data. 
The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), requiring only a small amount of use of the command line, which may be called from RStudio or Positron rather than a 'traditional' terminal. 
Further while the software is intended for use in the context of producing herbarium labels, it's basic functionality to resolve taxonomic names, and query environmental attributes may make it useful for various land management agencies which require these utilities for various research activities.

@@

# DATA AVAILABILITY STATEMENT
The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives). 
The package includes seven use-case vignettes (tutorials) available on a Github Pages site (https://sagesteppe.github.io/BarnebyLives/).
The first vignette *"Preparing to use BarnebyLives!"* shows how to set up an instance for a certain geographic area (domain).
The next vignette *"BarnebyLives! Running pipeline"* showcases the core functionality of the package for processing data entered on a spreadsheet. 
*"Printing herbarium labels and exporting a digital copy of data"* demonstrates how to export data in both digital and analog formats while *"Rendering a shipping manifest"* details how to produce a shipping manifest for gifting or transferring material to an herbarium. 
The remaining vignettes detail different aspects of label usage, such as including maps and logos in *"Labels with graphics"*, and customization for example the position of template elements *"Customizing a label template"*, and fonts in *"Customizing label fonts"*.

All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript.
